---
title: "Running NONMEM with nlmixr2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running NONMEM with nlmixr2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
if (!dir.exists("pk.turnover.emax3-nonmem")) {
  stop("not setup correctly")
}
if (file.exists("pk.turnover.emax3-nonmem/pk.turnover.emax3-rounding.qs")) {
  unlink("pk.turnover.emax3-nonmem/pk.turnover.emax3-rounding.qs")
}
if (file.exists("pk.turnover.emax4-nonmem/pk.turnover.emax4.qs")) {
  unlink("pk.turnover.emax4-nonmem/pk.turnover.emax4.qs")
}
```

```{r setup}
library(babelmixr2)
```

# Step 0: What do you need to do to have `nlmixr2` run `NONMEM` from a nlmixr2 model

To use `NONMEM` in nlmixr, you do not need to change your data or your
`nlmixr2` dataset. `babelmixr2` will do the heavy lifting here.

You do need to setup how to run `NONMEM`.  For many cases this is
easy; You simply have to figure out the command to run `NONMEM` (it is
often useful to use the full command path).  You can set it in
`options("babelmixr2.nonmem"="nmfe743")` or use
`nonmemControl(runCommand="nmfe743")`.  I prefer the `options()`
method since you only need to set it once.  This could also be a
function if you prefer (but I will not cover using the function here).

# Step 1: Run a `nlmixr2` in NONMEM

Lets take the classic warfarin example to start the comparison.

The model we use in the `nlmixr2` vignettes is:

```{r modelDef}
library(babelmixr2)
pk.turnover.emax3 <- function() {
  ini({
    tktr <- log(1)
    tka <- log(1)
    tcl <- log(0.1)
    tv <- log(10)
    ##
    eta.ktr ~ 1
    eta.ka ~ 1
    eta.cl ~ 2
    eta.v ~ 1
    prop.err <- 0.1
    pkadd.err <- 0.1
    ##
    temax <- logit(0.8)
    tec50 <- log(0.5)
    tkout <- log(0.05)
    te0 <- log(100)
    ##
    eta.emax ~ .5
    eta.ec50  ~ .5
    eta.kout ~ .5
    eta.e0 ~ .5
    ##
    pdadd.err <- 10
  })
  model({
    ktr <- exp(tktr + eta.ktr)
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v <- exp(tv + eta.v)
    emax = expit(temax+eta.emax)
    ec50 =  exp(tec50 + eta.ec50)
    kout = exp(tkout + eta.kout)
    e0 = exp(te0 + eta.e0)
    ##
    DCP = center/v
    PD=1-emax*DCP/(ec50+DCP)
    ##
    effect(0) = e0
    kin = e0*kout
    ##
    d/dt(depot) = -ktr * depot
    d/dt(gut) =  ktr * depot -ka * gut
    d/dt(center) =  ka * gut - cl / v * center
    d/dt(effect) = kin*PD -kout*effect
    ##
    cp = center / v
    cp ~ prop(prop.err) + add(pkadd.err)
    effect ~ add(pdadd.err) | pca
  })
}
```

Now you can run the `nlmixr2` model using `NONMEM` you simply can run
it directly:

```{r runRounding}
try(nlmixr(pk.turnover.emax3, nlmixr2data::warfarin, "nonmem",
           nonmemControl(readRounding=FALSE, modelName="pk.turnover.emax3")),
    silent=TRUE)
```

That this is the same way you would run an ordinary `nlmixr2`
model, but it is simply a new estimation method `"nonmem"` with a new controller (`nonmemControl()`)
to setup options for estimation.

A few options in the `nonmemControl()` here is `modelName` which helps
control the output directory of `NONMEM` (if not specified
`babelmixr2` tries to guess based on the model name based on the
input).

If you try this yourself, you see that `NONMEM` fails with rounding
errors. You could do the standard approach of changing `sigdig`,
`sigl`, `tol` etc, to get a successful `NONMEM` model convergence, of
course that is supported. But with `babelmixr2` you can *do more*.

# Optional Step 2: Recover a failed NONMEM run

One of the other approaches is to **ignore** the rounding errors that
have occurred and read into `nlmixr2` anyway:

```{r readRounding}
# Can still load the model to get information (possibly pipe) and create a new model
f <- nlmixr(pk.turnover.emax3, nlmixr2data::warfarin, "nonmem",
            nonmemControl(readRounding=TRUE, modelName="pk.turnover.emax3"))
```

You may see more work happening than you expected to need for an already
completed model.  When reading in a NONMEM model, `babelmixr2` grabs:

- `NONMEM`'s objective function value
- `NONMEM`'s covariance (if available)
- `NONMEM`'s optimization history
- `NONMEM`'s final parameter estimates (including the ETAs)
- `NONMEM`'s `PRED` and `IPRED` values (for validation purposes)

These are used to solve the ODEs *as if they came from an nlmixr2*
optimization procedure.

This means that you can compare the `IPRED` and `PRED` values of
`nlmixr2`/`rxode2` and *know immediately* if your model validates.

This is similar to the procedure Kyle Baron advocates for validating a
NONMEM model against a `mrgsolve` model (see
https://mrgsolve.org/blog/posts/2022-05-validate-translation/ and https://mrgsolve.org/blog/posts/2023-update-validation.html), 

The advantage of this method is that you need to simply write one model to
get a validated `roxde2`/`nlmixr2` model.

In this case you can see the validation when you print the fit object:

```{r printFit}
print(f)
```

Which shows the `preds` and `ipreds` match between `NONMEM` and
`nlmixr2` quite well.

# Optional Step 3: Use `nlmixr2` to help understand why `NONMEM` failed

Since it *is* a `nlmixr2` fit, you can do interesting things with this
fit that you couldn't do in `NONMEM` or even in another translator.
For example, if you wanted to add a covariance step you can with
`getVarCov()`:

```{r addCov}
getVarCov(f)
```

`nlmixr2` is more generous in what constitutes a covariance step.  The
`r,s` covariance matrix is the "most" successful covariance step for
`focei`, but the system will fall back to other methods if necessary.

While this covariance matrix is not `r,s`, and should be regarded with
caution, it can still give us some clues on why this things are not working in
`NONMEM`.

When examining the fit, you can see the shrinkage is high for `temax`,
`tktr` and `tka`, so they could be dropped, making things more likely
to converge in `NONMEM`.

# Optional Step 4: Use model piping to get a successful NONMEM run

If we use model piping to remove the parameters, the new run will
start at the last model's best estimates (saving a bunch of model development
time).

In this case, I specify the output directory `pk.turnover.emax4` with
the control and get the following:

```{r pipeToNewModel}
f2 <- f %>% model(ktr <- exp(tktr)) %>%
  model(ka <- exp(tka)) %>%
  model(emax = expit(temax)) %>%
  nlmixr(data=nlmixr2data::warfarin, est="nonmem",
         control=nonmemControl(readRounding=FALSE,
                               modelName="pk.turnover.emax4"))
```

You can see the `NONMEM` run is now successful and validates against
the `rxode2` model below:

```{r newModel}
f2
```

One thing to emphasize: unlike other translators, you will know immediately
if the translation is off because the model will not validate.  Hence
you can start this process with confidence - you will know
immediately if something is wrong.

This is related to [converting NONMEM to a nlmixr2 fit](https://nlmixr2.github.io/nonmem2rx/articles/convert-nlmixr2.html).
