---
title: "Exporting a model to PharmML"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exporting a model to PharmML}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(babelmixr2)
```

PharmML is an XML description of a pharmacometric model: the parameter model,
the structural model, the observation model, the trial design and the
estimation step, all in one self-describing document. Exporting to it makes an
`nlmixr2` model archivable and exchangeable without the receiving tool needing
to understand `rxode2` syntax.

Unlike the NONMEM and Monolix backends, this is a *writer*, not an estimation
method: `babelmixr2` does not run PharmML, so there is no `est="pharmml"`.

## Writing a model

```{r model}
one.cmt <- function() {
  ini({
    tka <- log(1.57); label("Ka")
    tcl <- log(2.72); label("Cl")
    tv  <- log(31.5); label("V")
    eta.ka ~ 0.6
    eta.cl ~ 0.3
    eta.v  ~ 0.1
    add.sd <- 0.7
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    vc <- exp(tv + eta.v)
    d/dt(depot)  <- -ka * depot
    d/dt(center) <-  ka * depot - cl / vc * center
    cp <- center / vc
    cp ~ add(add.sd)
  })
}
```

With no data, `as.pharmml()` writes just the model definition — a valid
PharmML description of the model, with no trial design or estimation step:

```{r modelOnly}
xml <- as.pharmml(one.cmt)
cat(substr(xml, 1, 600))
```

Pass the dataset as well and the document gains a `TrialDesign` and an
`EstimationStep`:

```{r withData}
xml <- as.pharmml(one.cmt, nlmixr2data::theo_sd)
```

To write it to disk, give a `file`. A PharmML document refers to its dataset
by path rather than embedding it, so the NONMEM-format CSV is written
alongside the model:

```{r write, eval = FALSE}
as.pharmml(one.cmt, nlmixr2data::theo_sd, file = "one.cmt.xml")
#> writes one.cmt.xml and one.cmt.csv
```

## Validation

Every document is checked against the PharmML schema before it is returned, so
a problem surfaces at the call rather than in whatever reads the file later.
The schemas are vendored in the package — the upstream host no longer serves
them — so validation is entirely offline.

```{r validate}
pharmmlValidate(xml)
pharmmlVersions()
```

`pharmmlControl(validate = FALSE)` turns the check off, though it costs only a
fraction of a second.

## What is and is not translated

PharmML's operator vocabulary is a close match for `rxode2`'s. `logit`,
`probit`, `logistic` (`expit`), `normcdf`, `log2`, `log10` and `gammaln` are
all native, so most model expressions map to a single operator.

Solved models are exported as PharmML *PK macros* rather than being expanded
to ODEs, so the structure the model was written in survives the round trip:

```{r linCmt}
solved <- function() {
  ini({
    tka <- log(1.57); tcl <- log(2.72); tv <- log(31.5)
    eta.cl ~ 0.3
    add.sd <- 0.7
  })
  model({
    ka <- exp(tka)
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv)
    linCmt() ~ add(add.sd)
  })
}
grepl("PKmacros", as.pharmml(solved))
```

Constructs PharmML cannot express raise an error naming the construct, rather
than emitting a document that looks plausible but is wrong. Currently out of
scope:

| Construct | Status |
|---|---|
| `add()`, `prop()`, `add() + prop()` | supported |
| `pow()`, `add() + pow()` | error |
| Non-normal residuals (`pois()`, `dbinom()`, `t()`, ordinal, `ll()`) | error |
| Inter-occasion variability | error |
| Mixture models (`mix()`) | error |
| Michaelis-Menten or transit absorption via `linCmt()` | error |
| Units | dropped, with a note in the description |

Categorical covariates are detected from the dataset: a character or factor
column becomes a PharmML `Categorical` covariate with a category per level,
and the column mapping records how the exported numeric codes correspond to
those levels. Without a dataset there is no type information to go on, so
covariates are written as continuous.

## Translating a single expression

`rxToPharmml()` is the expression-level translator, the PharmML counterpart of
`rxToNonmem()` and `rxToMonolix()`:

```{r expr}
cat(rxToPharmml("ka * depot"))
```

## Reading PharmML

There is no reader. `babelmixr2` writes PharmML but cannot import it, so the
output is validated against the schema and compared structurally against the
reference examples — it is never executed and read back. Importing PharmML
would be a separate package, in the way `nonmem2rx` and `monolix2rx` are
separate from the NONMEM and Monolix writers.
