---
title: "PopED"
author: "Matthew L. Fidler"
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")

# Setup the moodels from PopED's
# https://andrewhooker.github.io/PopED/articles/model_def_other_pkgs.html
library(babelmixr2)
library(PopED)
library(mrgsolve)
library(PKPDsim)
library(rxode2)

ff_analytic <- function(model_switch,xt,parameters,poped.db){
  with(as.list(parameters),{
    y=xt
    N = floor(xt/TAU)+1
    f=(DOSE/V)*(KA/(KA - CL/V)) *
      (exp(-CL/V * (xt - (N - 1) * TAU)) * (1 - exp(-N * CL/V * TAU))/(1 - exp(-CL/V * TAU)) -
         exp(-KA * (xt - (N - 1) * TAU)) * (1 - exp(-N * KA * TAU))/(1 - exp(-KA * TAU)))
    return(list( f=f,poped.db=poped.db))
  })
}


code <- '
$PARAM CL=3.75, V=72.8, KA=0.25
$CMT DEPOT CENT
$ODE
dxdt_DEPOT = -KA*DEPOT;
dxdt_CENT = KA*DEPOT - (CL/V)*CENT;
$TABLE double CP  = CENT/V;
$CAPTURE CP
'

moda <- mrgsolve::mcode("optim", code, atol=1e-8, rtol=1e-8,maxsteps=5000)

ff_ode_mrg <- function(model_switch, xt, p, poped.db){
  times_xt <- drop(xt)
  dose_times <- seq(from=0,to=max(times_xt),by=p[["TAU"]])
  time <- sort(unique(c(0,times_xt,dose_times)))
  is.dose <- time %in% dose_times

  data <-
    tibble::tibble(ID = 1,
                   time = time,
                   amt = ifelse(is.dose,p[["DOSE"]], 0),
                   cmt = ifelse(is.dose, 1, 0),
                   evid = cmt,
                   CL = p[["CL"]], V = p[["V"]], KA = p[["KA"]])

  out <- mrgsolve::mrgsim_q(moda, data=data)

  f <-  out$CP

  f <- f[match(times_xt,out$time)]

  return(list(f=matrix(f,ncol=1),poped.db=poped.db))

}

sfg <- function(x,a,bpop,b,bocc){
  parameters=c(
    KA=bpop[1]*exp(b[1]),
    CL=bpop[2]*exp(b[2]),
    V=bpop[3]*exp(b[3]),
    DOSE=a[1],
    TAU=a[2])
  return( parameters )
}

feps <- function(model_switch,xt,parameters,epsi,poped.db){
  f <- do.call(poped.db$model$ff_pointer,list(model_switch,xt,parameters,poped.db))[[1]]
  y = f*(1+epsi[,1])+epsi[,2]
  return(list(y=y,poped.db=poped.db))
}

poped_db_analytic <- create.poped.database(
  ff_fun =ff_analytic,
  fg_fun =sfg,
  fError_fun=feps,
  bpop=c(KA=0.25,CL=3.75,V=72.8),
  d=c(KA=0.09,CL=0.25^2,V=0.09),
  sigma=c(prop=0.04,add=0.0025),
  m=2,
  groupsize=20,
  xt=c( 1,2,8,240,245),
  minxt=c(0,0,0,240,240),
  maxxt=c(10,10,10,248,248),
  bUseGrouped_xt=1,
  a=cbind(DOSE=c(20,40),TAU=c(24,24)),
  maxa=c(DOSE=200,TAU=24),
  mina=c(DOSE=0,TAU=24))


poped_db_ode_mrg <- create.poped.database(poped_db_analytic,ff_fun = ff_ode_mrg)

pk1cmtoral <- PKPDsim::new_ode_model("pk_1cmt_oral") # take from library
ff_ode_pkpdsim <- function(model_switch, xt, p, poped.db){
    #Set up time points for the ODE
    times_xt <- drop(xt)  
    dose_times <- seq(from=0,to=max(times_xt),by=p[["TAU"]])
    times <- sort(unique(c(0,times_xt,dose_times)))

    N = length(dose_times)
    regimen = PKPDsim::new_regimen(amt=p[["DOSE"]],n=N,interval=p[["TAU"]])
    design <- PKPDsim::sim(
      ode = pk1cmtoral, 
      parameters = c(CL=p[["CL"]],V=p[["V"]],KA=p[["KA"]]), 
      regimen = regimen,
      only_obs = TRUE,
      t_obs = times,
      checks = FALSE,
      return_design = TRUE)
    tmp <- PKPDsim::sim_core(sim_object = design, ode = pk1cmtoral)
    f <- tmp$y
    m_tmp <- match(round(times_xt,digits = 6),tmp[,"t"])
    if(any(is.na(m_tmp))){
      stop("can't find time points in solution\n", 
           "try changing the digits argument in the match function")
    } 
    
    f <- f[m_tmp]
    return(list(f = f, poped.db = poped.db))
}

modrx <- rxode2::rxode2({
  d/dt(DEPOT) = -KA*DEPOT;
  d/dt(CENT) = KA*DEPOT - (CL/V)*CENT;
  CP=CENT/V;
})

ff_ode_rx <- function(model_switch, xt, p, poped.db){
  times_xt <- drop(xt)
  et(0,amt=p[["DOSE"]], ii=p[["TAU"]], until=max(times_xt)) %>%
    et(times_xt) -> data
  
  out <- rxode2::rxSolve(modrx, p, data, atol=1e-8, rtol=1e-8,maxsteps=5000,
                 returnType="data.frame")
  
  f <-  out$CP[match(times_xt,out$time)]
  
  return(list(f=matrix(f,ncol=1),poped.db=poped.db))
  
}

poped_db_ode_rx <- create.poped.database(poped_db_analytic,ff_fun = ff_ode_rx)

poped_db_ode_pkpdsim <- create.poped.database(poped_db_analytic,ff_fun = ff_ode_pkpdsim)
```

## Introduction -- using babelmixr2 with PopED

`babelmixr2` now introduces a new method taking `rxode2`/`nlmixr2`
models and optimal design using `PopED`.

As in the [PopED vignette comparing ODE
solvers](https://andrewhooker.github.io/PopED/articles/model_def_other_pkgs.html#speed-of-fim-computation)
(and their speeds), this section will:

-   take the model described and adapt it in two different `rxode2`
    model functions, the solved and ode cases
-   compare these examples to the fastest solvers in the PopED vignette
    (`mrgsolve` and `analytic`)

## babelmixr2 Ode solution

```{r babelmixr2_ode}
library(babelmixr2)
library(PopED)

e <- et(amt=1, ii=24, until=250) %>%
  et(list(c(0, 10),
          c(0, 10),
          c(0, 10),
          c(240, 248),
          c(240, 248))) %>%
  dplyr::mutate(time =c(0, 1,2,8,240,245))

# model
f <- function() {
  ini({
    tV <- 72.8
    tKA <- 0.25
    tCL <- 3.75
    Favail <- fix(0.9)
    eta.ka ~ 0.09
    eta.cl ~ 0.25 ^ 2
    eta.v ~ 0.09
    prop.sd <- sqrt(0.04)
    add.sd <- fix(sqrt(5e-6))
  })
  model({
    ka <- tKA * exp(eta.ka)
    v <- tV * exp(eta.v)
    cl <- tCL * exp(eta.cl)
    d/dt(depot) <- -ka * depot
    d/dt(central) <- ka * depot - cl / v * central
    cp <- central / v
    f(depot) <- DOSE * Favail
    cp ~ add(add.sd) + prop(prop.sd)
  })
}

poped_db_ode_babelmixr2 <- nlmixr(f, e,
                                  popedControl(a=list(c(DOSE=20),
                                                      c(DOSE=40)),
                                               maxa=c(DOSE=200),
                                               mina=c(DOSE=0)))
f2 <- function() {
  ini({
    tV <- 72.8
    tKA <- 0.25
    tCL <- 3.75
    Favail <- fix(0.9)
    eta.ka ~ 0.09
    eta.cl ~ 0.25 ^ 2
    eta.v ~ 0.09
    prop.sd <- sqrt(0.04)
    add.sd <- fix(sqrt(5e-6))
  })
  model({
    ka <- tKA * exp(eta.ka)
    v <- tV * exp(eta.v)
    cl <- tCL * exp(eta.cl)
    cp <- linCmt()
    f(depot) <- DOSE * Favail
    cp ~ add(add.sd) + prop(prop.sd)
  })
}

poped_db_analytic_babelmixr2 <- nlmixr(f, e,
                                       popedControl(a=list(c(DOSE=20),
                                                           c(DOSE=40)),
                                                    maxa=c(DOSE=200),
                                                    mina=c(DOSE=0)))


library(ggplot2)
library(microbenchmark)

compare <- microbenchmark(
  evaluate_design(poped_db_analytic),
  evaluate_design(poped_db_analytic_babelmixr2),
  evaluate_design(poped_db_ode_babelmixr2),
  evaluate_design(poped_db_ode_rx),
  evaluate_design(poped_db_ode_mrg),
  evaluate_design(poped_db_ode_pkpdsim),
  times = 100L)

autoplot(compare)

```

Note that the `babelmixr2` ode solver is the fastest ode solver in this
comparison. Among other things, this is because the model is loaded into
memory and does not need to be setup each time. (As benchmarks, the
`mrgsolve`, `PKPDsim` and `rxode2` implementation on the `PopED`'s
website is included).

Note that the speed of all the tools are reasonable. In my opinion, the
benefit of the `babelmixr2` interface to `PopED` is the simplicity of
using `nlmixr2` / `rxode2` functional models or fits directly in `PopED`
without relying on conversions.

The interface is a bit different than the traditional `PopED` interface,
and requires a design data-set as well as a `popedControl()` to setup a
`PopED` database to run all of the `PopED` tasks. This is because
traditionally `nlmixr2` takes a dataset, "estimation" method and
controls to change estimation method options.

This allows the same type of paradigm to be applied to `PopED`. In my
estimation, this should allow easy translation between the systems and
more frequent use of optimal design in designingdd clinical trials.
