---
title: "PopED"
author: "Matthew L. Fidler"
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")

# Setup the moodels from PopED's
# https://andrewhooker.github.io/PopED/articles/model_def_other_pkgs.html
library(babelmixr2)

library(PopED)

library(mrgsolve)

library(PKPDsim)

library(rxode2)

ff_analytic <- function(model_switch,xt,parameters,poped.db){
  with(as.list(parameters),{
    y=xt
    N = floor(xt/TAU)+1
    f=(DOSE/V)*(KA/(KA - CL/V)) *
      (exp(-CL/V * (xt - (N - 1) * TAU)) * (1 - exp(-N * CL/V * TAU))/(1 - exp(-CL/V * TAU)) -
         exp(-KA * (xt - (N - 1) * TAU)) * (1 - exp(-N * KA * TAU))/(1 - exp(-KA * TAU)))
    return(list( f=f,poped.db=poped.db))
  })
}


code <- '
$PARAM CL=3.75, V=72.8, KA=0.25
$CMT DEPOT CENT
$ODE
dxdt_DEPOT = -KA*DEPOT;
dxdt_CENT = KA*DEPOT - (CL/V)*CENT;
$TABLE double CP  = CENT/V;
$CAPTURE CP
'

moda <- mrgsolve::mcode("optim", code, atol=1e-8, rtol=1e-8,maxsteps=5000)

ff_ode_mrg <- function(model_switch, xt, p, poped.db){
  times_xt <- drop(xt)
  dose_times <- seq(from=0,to=max(times_xt),by=p[["TAU"]])
  time <- sort(unique(c(0,times_xt,dose_times)))
  is.dose <- time %in% dose_times

  data <-
    tibble::tibble(ID = 1,
                   time = time,
                   amt = ifelse(is.dose,p[["DOSE"]], 0),
                   cmt = ifelse(is.dose, 1, 0),
                   evid = cmt,
                   CL = p[["CL"]], V = p[["V"]], KA = p[["KA"]])

  out <- mrgsolve::mrgsim_q(moda, data=data)

  f <-  out$CP

  f <- f[match(times_xt,out$time)]

  return(list(f=matrix(f,ncol=1),poped.db=poped.db))

}

sfg <- function(x,a,bpop,b,bocc){
  parameters=c(
    KA=bpop[1]*exp(b[1]),
    CL=bpop[2]*exp(b[2]),
    V=bpop[3]*exp(b[3]),
    DOSE=a[1],
    TAU=a[2])
  return( parameters )
}

feps <- function(model_switch,xt,parameters,epsi,poped.db){
  f <- do.call(poped.db$model$ff_pointer,list(model_switch,xt,parameters,poped.db))[[1]]
  y = f*(1+epsi[,1])+epsi[,2]
  return(list(y=y,poped.db=poped.db))
}

poped_db_analytic <- create.poped.database(
  ff_fun =ff_analytic,
  fg_fun =sfg,
  fError_fun=feps,
  bpop=c(KA=0.25,CL=3.75,V=72.8),
  d=c(KA=0.09,CL=0.25^2,V=0.09),
  sigma=c(prop=0.04,add=0.0025),
  m=2,
  groupsize=20,
  xt=c( 1,2,8,240,245),
  minxt=c(0,0,0,240,240),
  maxxt=c(10,10,10,248,248),
  bUseGrouped_xt=1,
  a=cbind(DOSE=c(20,40),TAU=c(24,24)),
  maxa=c(DOSE=200,TAU=24),
  mina=c(DOSE=0,TAU=24))


poped_db_ode_mrg <- create.poped.database(poped_db_analytic,ff_fun = ff_ode_mrg)

pk1cmtoral <- PKPDsim::new_ode_model("pk_1cmt_oral") # take from library
ff_ode_pkpdsim <- function(model_switch, xt, p, poped.db){
    #Set up time points for the ODE
    times_xt <- drop(xt)
    dose_times <- seq(from=0,to=max(times_xt),by=p[["TAU"]])
    times <- sort(unique(c(0,times_xt,dose_times)))
  #
    N = length(dose_times)
    regimen = PKPDsim::new_regimen(amt=p[["DOSE"]],n=N,interval=p[["TAU"]])
    design <- PKPDsim::sim(
      ode = pk1cmtoral,
      parameters = c(CL=p[["CL"]],V=p[["V"]],KA=p[["KA"]]),
      regimen = regimen,
      only_obs = TRUE,
      t_obs = times,
      checks = FALSE,
      return_design = TRUE)
    tmp <- PKPDsim::sim_core(sim_object = design, ode = pk1cmtoral)
    f <- tmp$y
    m_tmp <- match(round(times_xt,digits = 6),tmp[,"t"])
    if(any(is.na(m_tmp))){
      stop("can't find time points in solution\n",
           "try changing the digits argument in the match function")
    }
   #
    f <- f[m_tmp]
    return(list(f = f, poped.db = poped.db))
}

modrx <- rxode2::rxode2({
  d/dt(DEPOT) = -KA*DEPOT;
  d/dt(CENT) = KA*DEPOT - (CL/V)*CENT;
  CP=CENT/V;
})

ff_ode_rx <- function(model_switch, xt, p, poped.db){
  times_xt <- drop(xt)
  et(0,amt=p[["DOSE"]], ii=p[["TAU"]], until=max(times_xt)) %>%
    et(times_xt) -> data

  out <- rxode2::rxSolve(modrx, p, data, atol=1e-8, rtol=1e-8,maxsteps=5000,
                 returnType="data.frame")

  f <-  out$CP[match(times_xt,out$time)]

  return(list(f=matrix(f,ncol=1),poped.db=poped.db))

}

poped_db_ode_rx <- create.poped.database(poped_db_analytic,ff_fun = ff_ode_rx)

poped_db_ode_pkpdsim <- create.poped.database(poped_db_analytic,ff_fun = ff_ode_pkpdsim)
```

## Introduction -- using babelmixr2 with PopED

`babelmixr2` now introduces a new method that takes `rxode2`/`nlmixr2`
models converts them to a `PopED` database to help with optimal design.

As in the [PopED vignette comparing ODE
solvers](https://andrewhooker.github.io/PopED/articles/model_def_other_pkgs.html#speed-of-fim-computation)
(and their speeds), this section will:

-   take the model described and adapt it in two different `rxode2`
    model functions, the solved and ode cases (this is done by the
    `nlmixr()` call which creates a `PopED` database)

-   compare these examples to the pharmacometric solvers in the PopED
    vignette (`mrgsolve`, original `rxode2` and `PKPDsim`)

## babelmixr2 ODE solution

The first step of a design using babelmixr2 is to tell babelmixr2
about the design being optimized.  This is a bit different than what
is done in `PopED` directly.  Below I am using the `et()` function to
create the event table like a typical `rxode2` simulation, but it is
used to specify the study design:

```{r babelmixr2-events}

library(babelmixr2)
library(PopED)

e <- et(amt=1, ii=24, until=250) %>%
  et(time=c(1,2,8,240,245)) %>%
  as.data.frame() %>%
  dplyr::mutate(low=c(NA_real_, 0, 0, 0, 240, 240),
                high=c(NA_real_, 10, 10, 10, 248, 248))


print(e)
```

### PopED/babelmixr2 event table description

Here note that `time` is the design times for the `PopED` designs,
they can include dosing; only observations are considered the
time-points.  They become the `xt` parameter in the `PopED` database
(excluding the doses).

We also build on the structure of the `rxode2` event table with
simulations.  In simulations the sampling windows cause random times
to be generated inside the sampling windows.  For this reason, the
last line of code fixes the times to where we want to have the
multiple endpoint design.

Therefore, in this dataset the `low` becomes `minxt` and `high` becomes `maxxt`.

We chose these because they build on what is already know from
`nlmixr2` and used in and do not require any extra coding.

Other things you may have to include in your `PopED` model data frame are:

- `dvid` which gives the integer of the model endpoint measured (like
  `rxode2` but has to be an integer).  This becomes `model_switch` in
  the `PopED` dataset.

- `G_xt` which is the `PopED` grouping variable; This will be put into
  the `PopED` database as `G_xt`

- `id` becomes an ID for a design (which you can use as a covariate to
  pool different designs or different regimens for optimal design).

### Getting PopED functions from `nlmixr2`/`rxode2` ui function

Once the design is setup, we need to specify a model.  It is easy to
specify the model using the `nlmixr2`/`rxode2` function/ui below:

```{r babelmixr2-model}
# model
f <- function() {
  ini({
    tKA <- 0.25
    tCL <- 3.75
    tV <- 72.8
    Favail <- fix(0.9)
    eta.ka ~ 0.09
    eta.cl ~ 0.25 ^ 2
    eta.v ~ 0.09
    prop.sd <- sqrt(0.04)
    add.sd <- sqrt(0.0025)
  })
  model({
    ka <- tKA * exp(eta.ka)
    v <- tV * exp(eta.v)
    cl <- tCL * exp(eta.cl)
    d/dt(depot) <- -ka * depot
    d/dt(central) <- ka * depot - cl / v * central
    cp <- central / v
    f(depot) <- DOSE * Favail
    cp ~ add(add.sd) + prop(prop.sd)
  })
}

f <- f() # compile/check nlmixr2/rxode2 model


# Create a PopED database for `nlmixr2`:
poped_db_ode_babelmixr2 <- nlmixr(f, e, "poped",
                                  popedControl(a=list(c(DOSE=20),
                                                      c(DOSE=40)),
                                               maxa=c(DOSE=200),
                                               mina=c(DOSE=0)))
```

Note when creating a `PopED` database with a model and a design event
table, many of the `PopED` database components are generated for you.

These are not things that are hidden, but things you can access
directly from the model or even from the compiled `ui`. Much of the
other options for optimal design can be specified with the
`popedControl()` function.

This can help you understand what `babelmixr2` is doing, we will show
what is being added:

#### PopED's `ff_fun` from babelmixr2

This is the function that is run to generate the predictions:

```{r}
# The ff_fun can be retrieved from the ui with f$popedFfFun
f$popedFfFun
```

Some things to note in this function:

- The model changes based on the number of time-points requested.  In
  this case it is `5` since there were `5` design points in the design above.

- There are some `babelmixr2` specific functions here:

  - `babelmixr2::popedMultipleEndpointParam`, which indexes the time
    input and model_input to make sure the input matches as requested
    (in many PopED functions they use `match.time` and this is a bit
    similar).

  - `.popedRxRunSetup`/`.popedRxRunFullSetupMe` which runs the rxode2
    setup including loading the data and model into memory (and is a
    bit different depending on the number of time points you are
    using)

  - `.popedSolveIdME`/`.popedSolveIdME2` which solves the rxode2 model and uses the
    indexes to give the solve used in the model.

#### rxode2 models generated from `babelmixr2`

By describing this, you can also see that there are 2 `rxode2` models
generated for the `PopED` database.  You can see these inside of the
PopED database as well.

The first model uses model times to solve for arbitrary times based on design:

```{r showModelsMT}
summary(poped_db_ode_babelmixr2$babelmixr2$modelMT)
```

You can also see the model used for solving scenarios with a number of
time points greater than the design specification:


```{r showModelsF}
summary(poped_db_ode_babelmixr2$babelmixr2$modelF)
```

You can also see that the models are identical with the exception of
requesting modeled times.  You can see the base/core `rxode2` model
form the UI here:

```{r baseModelForPop}
f$popedRxmodelBase
```

#### PopED's `fg_fun`

`babelmixr2` also generates `PopED`s `fg_fun`, which translates
covariates and parameters into the parameters required in the `ff_fun`
and used in solving the `rxode2` model.

```{r fg_fun}
# You can see the PopED fg_fun from the model UI with
# f$popedFgFun:
f$popedFgFun
```

#### PopED's error function `fError_fun`

You can see the `babelmixr2` generated error function as well with:

```{r popedErr}
f$popedFErrorFun
```

One really important note to keep in mind is that `PopED` works with
variances instead of standard deviations (which is a key difference
between `nlmixr2` and `PopED`).

This means that the exported model from `babelmixr2` operates on
variances instead of standard deviations and care must be taken in
using these values to not misinterpret the two.

The export also tries to flag this to make it easier to remember.

#### Other parameters generated by `PopED`

```{r others}
f$popedBpop # PopED bpop
f$popedNotfixedBpop # PopED notfixed_bpop
f$popedD # PopED d
f$popedNotfixedD # PopED notfixed_d
f$popedCovd # PopED covd
f$popedNotfixedCovd # PopED notfixed_covd
f$popedSigma # PopED sigma (variance is exported, not SD)
f$popedNotfixedSigma # PopED notfixed_sigma
```

The rest of the parameters are generated in conjunction with the
`popedControl()`.

## linear compartment models in `babelmixr2`

You can also specify the models using the `linCmt()` solutions as
below:

```{r lincmt}
f2 <- function() {
  ini({
    tV <- 72.8
    tKA <- 0.25
    tCL <- 3.75
    Favail <- fix(0.9)
    eta.ka ~ 0.09
    eta.cl ~ 0.25 ^ 2
    eta.v ~ 0.09
    prop.sd <- sqrt(0.04)
    add.sd <- fix(sqrt(5e-6))
  })
  model({
    ka <- tKA * exp(eta.ka)
    v <- tV * exp(eta.v)
    cl <- tCL * exp(eta.cl)
    cp <- linCmt()
    f(depot) <- DOSE
    cp ~ add(add.sd) + prop(prop.sd)
  })
}

poped_db_analytic_babelmixr2 <- nlmixr(f, e,
                                       popedControl(a=list(c(DOSE=20),
                                                           c(DOSE=40)),
                                                    maxa=c(DOSE=200),
                                                    mina=c(DOSE=0)))
```


## Comparing method to the speed of other methods

```{r compare}
library(ggplot2)
library(microbenchmark)

compare <- microbenchmark(
  evaluate_design(poped_db_analytic),
  evaluate_design(poped_db_analytic_babelmixr2),
  evaluate_design(poped_db_ode_babelmixr2),
  evaluate_design(poped_db_ode_mrg),
  evaluate_design(poped_db_ode_pkpdsim),
  evaluate_design(poped_db_ode_rx),
  times = 100L)


autoplot(compare) + theme_bw()
```

Note that the `babelmixr2` ode solution is the fastest ode solver in this
comparison. Among other things, this is because the model is loaded into
memory and does not need to be setup each time. (As benchmarks, the
`mrgsolve`, and `PKPDsim` implementations on the `PopED`'s
website are included).

Also to me, the speed of all the tools are reasonable. In my opinion,
the benefit of the `babelmixr2` interface to `PopED` is the simplicity
of using `nlmixr2` / `rxode2` functional models or fits directly in
`PopED` without relying on conversions.

The interface is a bit different than the traditional `PopED` interface,
and requires a design data-set as well as a `popedControl()` to setup a
`PopED` database to run all of the `PopED` tasks. This is because
traditionally `nlmixr2` takes a dataset, "estimation" method and
controls to change estimation method options.

`babelmixr2` adopts the same paradigm of model, data, control to be
applied to `PopED`. This should allow easy translation between the
systems. With easier translation, hopefully optimal design in clinical
trials will be easier to achieve.

# Key notes to keep in mind

- `babelmixr2` loads models into memory and needs to keep track of
  which model is loaded. To help this you need to use
  `babel.poped.database` in place of `create.poped.database` when
  modifying babelmixr2 generated `PopED` databases.  If this isn't
  done, there is a chance that the model loaded will not be the
  expected loaded model and may either crash R or possibly give incorrect
  results.

- `babelmixr2` translates all error components to variances instead of
  the standard deviations in the `nlmixr2`/`rxode2` model

- When there are covariances in the `omega` specification, they will
  be identified as `D[#,#]` in the `PopED` output. To see what these
  numbers refer to it is helpful to see the name translations with
  `$popedD`.

- Depending on your options, `babelmixr2` may literally fix the model
  components, which means indexes may be different than you
  expect. The best way to get the correct index is use the
  `babelmixr2` function `babelBpopIdx()` which is useful for using
  `PopED`

- `babelmixr2` uses model times in creating `PopED` databases;
  therefore models with modeling times in them cannot be used in this translation

- `babelmixr2` does not yet support inter-occasion variability models.

## Where to find more examples

The examples from `PopED` have been converted to work with `babelmixr2` and are available in the package and on [GitHub](https://github.com/nlmixr2/babelmixr2/tree/main/inst/poped)
