---
title: "Running Monolix"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo=TRUE
)
```

```{r setup, include=FALSE}
if (!dir.exists("pk.turnover.emax3-monolix")) {
  stop("not setup correctly")
} else {
  if (file.exists("pk.turnover.emax3-monolix/nlmixr.qs")) {
    unlink("pk.turnover.emax3-monolix/nlmixr.qs")
  }
}
library(rxode2)
library(dplyr)
library(babelmixr2)
```


# Step 0: What do you need to do to have `nlmixr2` run `Monolix` from a nlmixr2 model

To use `Monolix` with nlmixr2, you do not need to change your data or your
`nlmixr2` dataset. `babelmixr2` will do the heavy lifting here.

You do need to setup how to run `Monolix`.  If you have setup the
`lixoftConnectors` package from Monolix, no further setup is needed.
Instead if you run `Monolix` from the command line for grid processing
(for example) you can figure out the command to run `Monolix` (it is
often useful to use the full command path and set it in the options,
ie `options("babelmixr2.monolix"="monolix")` or use
`monolixControl(runCommand="monolix")`.  If needed, I prefer the
`options()` method since you only need to set it once.  This could
also be a function if you prefer (but I will not cover using the
function here).

# Step 1: Run a `nlmixr2` in Monolix

Lets take the classic warfarin example. The model we use in the `nlmixr2` vignettes is:

```{r defineModel}
pk.turnover.emax3 <- function() {
  ini({
    tktr <- log(1)
    tka <- log(1)
    tcl <- log(0.1)
    tv <- log(10)
    ##
    eta.ktr ~ 1
    eta.ka ~ 1
    eta.cl ~ 2
    eta.v ~ 1
    prop.err <- 0.1
    pkadd.err <- 0.1
    ##
    temax <- logit(0.8)
    tec50 <- log(0.5)
    tkout <- log(0.05)
    te0 <- log(100)
    ##
    eta.emax ~ .5
    eta.ec50  ~ .5
    eta.kout ~ .5
    eta.e0 ~ .5
    ##
    pdadd.err <- 10
  })
  model({
    ktr <- exp(tktr + eta.ktr)
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v <- exp(tv + eta.v)
    emax = expit(temax+eta.emax)
    ec50 =  exp(tec50 + eta.ec50)
    kout = exp(tkout + eta.kout)
    e0 = exp(te0 + eta.e0)
    ##
    DCP = center/v
    PD=1-emax*DCP/(ec50+DCP)
    ##
    effect(0) = e0
    kin = e0*kout
    ##
    d/dt(depot) = -ktr * depot
    d/dt(gut) =  ktr * depot -ka * gut
    d/dt(center) =  ka * gut - cl / v * center
    d/dt(effect) = kin*PD -kout*effect
    ##
    cp = center / v
    cp ~ prop(prop.err) + add(pkadd.err)
    effect ~ add(pdadd.err) | pca
  })
}
```
Once monolix is run, you can run the `nlmixr2` model using `Monolix` as if it is new estimation method:

```{r runModel}
fit <- nlmixr(pk.turnover.emax3, nlmixr2data::warfarin, "monolix",
                              monolixControl(modelName="pk.turnover.emax3"))
```

This fit issues an informational tidbit -

- monolix parameter history needs exported charts, please export charts

This will automatically be generated as well when `lixoftConnectors`
package is generated and you have a recent version of Monolix.  If you
don't have that information then the important parameter history plots will 
not be imported and you cannot see those plots.

Just like with the NONMEM translation, the `monolixControl()` has
`modelName` which helps control the output directory of `Monolix` (if
not specified `babelmixr2` tries to guess based on the model name
based on the input).

Printing this out this `nlmixr2` fit you see:

```{r displayFit}
fit
```

Of particular interest is the comparison between Monolix predictions
and nlmixr predictions. In this case, I believe that these also imply
the models are predicting the same thing.  Note that the model
predictions are not as close as they were with [NONMEM](https://nlmixr2.github.io/babelmixr2/articles/running-nonmem.html) because Monolix
does not use the `lsoda` ODE solver.  Hence this small deviation is
expected, but still gives a validated Monolix model.

# Optional Step 2: Add conditional weighted residuals/focei objf to Monolix

As in the case of `NONMEM`, this gives some things that are not
available to Monolix, like adding conditional weighted residuals:

```{r addCwres}
fit <- addCwres(fit)
```


Which will add nlmixr's CWRES as well as adding the nlmixr2 `FOCEi` objective function

Because you now have an objective function compared based on the same
assumptions, you could compare the performance of Monolix and NONMEM
based on objective function.

To be fair, objective function values must always be used with
caution.  How the model performs and predicts the data is far more
valuable.

# Optional Step 3: Use nlmixr2 for vpc, reporting, etc.

Also since it is a `nlmixr2` object it would be easy to perform a VPC
too:

```{r vpc}
v1s <- vpcPlot(fit, show=list(obs_dv=TRUE), scales="free_y") +
  ylab("Warfarin Cp [mg/L] or PCA") +
  xlab("Time [h]")

v2s <- vpcPlot(fit, show=list(obs_dv=TRUE), pred_corr = TRUE, scales="free_y") +
  ylab("Prediction Corrected Warfarin Cp [mg/L] or PCA") +
  xlab("Time [h]")

v1s

v2s
```

# Notes about Monolix data translation

The input dataset expected to be compatible with `rxode2` or
`nlmixr2`.  This dataset is then converted to Monolix format:

- The combination of `CMT` and Dose type creates a unique `ADM` variable. 

- The `ADM` definition is saved in the monolix model file 

- `babelmixr2` creates a macro describing the compartment, ie `compartment(cmt=#, amount=stateName)`

- `babelmixr2` also creates a macro for each type of dosing:

   - Bolus/infusion uses `depot()` and adds modeled lag time (`Tlag`) or bioavailability (`p`) if specified
   
   - Modeled rate uses `depot()` with `Tk0=amtDose/rate`. `babelmixr2`
     also adds modeled lag time (`Tlag`) or bioavailability (`p`) if
     specified
   
   - Modeled duration uses `depot()` with `Tk0=dur`, also add adds
     modeled lag time (`Tlag`) or bioavailability (`p`) if specified
     Turning off a compartment uses empty macro
   
